---
title: "MTH6139 Time Series" 
subtitle: "Coursework 1 _ 231180728" 
author: "Jixiu WU" 
date: "Spring term 2025" 
output: 
  html_document:
    toc: true
    toc_float: true
    theme: spacelab 
    highlight: tango
---



```{r, echo=FALSE}
# This code will display the QMUL logo at the top right of the page
# Do not change this code
htmltools::img(src = knitr::image_uri("images/QMlogo.png"),
               alt = 'logo',
               style = 'position:absolute; top:0; right:0; padding:10px; width:20%;')
```


# 1. Introduction

This project shows the monthly Consumer Price Index of Transportation in U.S. City average for all urban consumers from January 1947 to February 2026.
It measures how the price of transportation change over time.

The purpose of the project is to analyse the monthly Consumer Price Index time series.
The objective is to identify trend and seasonal patterns, then build a time series model using Prophet.

Using Prophet is because Prophet could capture both trend and seasonality in time series data in a simple way.




# 2. Data

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(prophet)
library(readr)

transport_cpi_rawdata <- read_csv("data/CPITRNSL.csv")
transport_df <- data.frame(
    ds = as.Date(transport_cpi_rawdata$observation_date),
    y = transport_cpi_rawdata$CPITRNSL
)
head(transport_df)
model = prophet(transport_df)
future = make_future_dataframe(model, 8, freq = "month")
forecast = predict(model, future)
plot(model, forecast)
```




# 3.Trend

The overall chart shows a clear upward trend.
From 1947 to 1974, there was a slow increase. A strong upward trend emerged starting specifically in 1974.
Despite the impact of the COVID-19 pandemic, the trend rise continually with notable fluctuations between 2019 and 2023.
After 2024, prices have risen sharply.




# 4.Seasonality

```{r echo=FALSE, warning=FALSE}
prophet_plot_components(model, forecast)
```
The seasonal component shows some variations within a year. There are several peaks and dips at different times of the year.
However, the magnitude of changes is relatively small compared to the overall trend.




# 5.Forecasting Results

The forecasting result shows that the monthly Consumer Price Index will continue increasing in the future.
This follows the strong upward trend in the data.
The uncertainty intervals show some variations around predictions.
Therefore, there are some uncertainty in the forecast. 




# 6.Summary

ln summary, this project analyses the monthly Consumer Price Index of Transportation time series.
This data shows a clear upward trend from 1947 to 2026. Seasonal fluctuations were present but minor relative to the trend. The forecast shows that prices of transportation will probably still rise in the future, ebven there might be some short-term fluctuations.
However, the results shoud be treated with caution, because unexpected events could affect future changes.


# References

- RStudio reference: <https://rmarkdown.rstudio.com/lesson-1.html>
- Data reference: <https://fred.stlouisfed.org/series/CPITRNSL>
